Extrinsic Type Inference (DRAFT 2025-06-09)

### 2025-07-11 extrinsic_type_inference_paper/paper.tex
update
add factorization soundness
add fresh check
update diff case
udpate overview examples

\examp{
  \J{def foo = ? f => (f(<succ> <succ> <zero> @), f(<true> @))}
}

@@

\Par{Structural Generalization} 
Functions on lists can be generic, in that they may be independent of the item values, as in \code{repeat}.
Additionally, they may factor out the details that depend on item values, as in \code{sort}.

@@

\noindent
Having many different specializations burdens the client of
the generic function to know many different names for roughly the same behavior and can obscure the meaning
of code at the call site. 
Untyped languages or extrinsically typed languages can bypass this rigidity found in intrinsically typed by specializing in multiple ways.

\examp{
  ...
  \\
  \J{def genCmp = [a,b => } 
  \\
  \I \J{a |> (}
  \\
  \I\I \J{[zero;@ => scalarCmp(a,b)]}
  \\
  \I\I \J{[succ;n => scalarCmp(a,b)]}
  \\
  \I\I \J{[nil;@ => lexicoCmp(a,b)]}
  \\
  \I\I \J{[cons;(x,xs) => lexicoCmp(a,b)]}
  \\
  \I \J{)}
  \\
  \J{] in} 
  \\
  \J{def genSort : (TOP}
  \\
  \I \J{\& (}\textbf{List}(\textbf{Nat}) \J{ -> } \textbf{List}(\textbf{Nat})
  \\
  \I \J{\& (}\textbf{List}(\textbf{List}(\textbf{nat})) \J{ -> } \textbf{List}(\textbf{List}(\textbf{Nat}))
  \\
  \J{) = sort(genCmp) in}
  \\
  \J{...} 
}

\noindent
By combining the scalar and lexicographic comparison procedures into a single comparison
procedure, a generalized sort function can be used to work on either scalars or vectors,
thereby removing the burden of choosing the proper comparison function at the call site.

@@

    \inferrule {
      \TODO{modify with case for non-recursive parameter type}
      \\\\
      \Theta, \Delta, \Gamma \entails e \hastype \tau \given  \Theta', \Delta'
      \\
      \Theta', \Delta' \entails \tau \subtypes \alpha \rightarrow \Pi
      \\\\
      \alpha \downarrow \Pi
      \fallingdotseq 
      \alpha' \uparrow \Pi'
      \\
      \textbf{pack}^-(\textbf{ftv}(\Delta')\ \alpha' \entails \Pi') = \tau'
      \\\\
      \factorsinto \J{LFP[} \alpha' \J{]}\tau' \subtypes \J{left} \mapsto \tau_l
      \\
      \factorsinto \J{LFP[} \alpha' \J{]}\tau' \subtypes \J{right} \mapsto \tau_r
    } {
      \Theta, \Delta, \Gamma \entails \J{loop(}e\J{)} 
      \hastype 
      \tau_l \J{->} \tau_r
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

\begin{definition} 
  \label{def:proof_subtyping_difference_intro}
  \emph{proof subtyping (difference introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \subtypes \tau_r \typdiff \eta \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \Vvdash \eta
      \\
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r \given \Theta', \Delta' 
      \\
      \nexists \Theta', \Delta' \qua  
      \Theta, \Delta \entails 
      \tau \subtypes \eta \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r \typdiff \eta
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

\begin{requirement}
  \label{req:schema_sequence_duality_correspondence}
  \emph{schema sequence duality correspondence}
  \tiny
  \nopad
  \begin{mathpar}
    \inferrule {
      \alpha_\nu \downarrow \Pi_\nu \fallingdotseq \alpha_\mu \uparrow \Pi_\mu
      \\
      \delta \satisfies \Delta
    } {
      (
      \forall \Theta^\dagger, \Delta^\dagger, \tau^\dagger \qua
      (\Theta^\dagger,\Delta^\dagger, \tau^\dagger) \in \Pi_\nu \implies
      \exists \delta^\dagger \qua \textbf{dom}(\delta^\dagger) \subseteq \Theta^\dagger \up
      \\\\
      (
      \forall \delta' \qua
      \textbf{dom}(\delta') \cap \textbf{ftv}(\Delta) = \epsilon \implies
      % \delta\ \alpha_\nu \slash \J{BOT} \oplus \delta' \oplus \delta^\dagger \satisfies \Delta^\dagger \implies
      % \delta\ \alpha_\nu \slash \J{BOT} \oplus \delta' \oplus \delta^\dagger \satisfies e_f \hastype \tau^\dagger
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies \Delta \cup \Delta^\dagger \implies
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies e_f \hastype \tau^\dagger
      )
      )
      \\\\
      \iff
      \\\\
      (\forall e_a, e_r \qua
      (
      \exists \Theta^\dagger, \Delta^\dagger, \tau^\dagger \qua
      (\Theta^\dagger,\Delta^\dagger, \tau^\dagger) \in \Pi_\mu \up
      (\forall \delta^\dagger \qua \textbf{dom}(\delta^\dagger) \subseteq \Theta^\dagger \implies
      \\\\
      \exists \delta' \qua
      \textbf{dom}(\delta') \cap \textbf{ftv}(\Delta) = \epsilon \up
      % \delta\ \alpha_\mu \slash \J{TOP} \oplus \delta' \oplus \delta^\dagger \satisfies \Delta^\dagger \up
      % \delta\ \alpha_\mu \slash \J{TOP} \oplus \delta' \oplus \delta^\dagger \satisfies (e_a \J{,} e_r) \hastype \tau^\dagger 
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies \Delta \cup \Delta^\dagger \up
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies (e_a \J{,} e_r) \hastype \tau^\dagger 
      )
      )
      \implies
      e_r \cong e_f\J{(}e_a\J{)}
      )
    }
  \end{mathpar}
\end{requirement}

@@

\TODO{add requirements on tidying}

\TODO{add requirements on factoring}

@@

>>>>

\examp{
  \J{def foo = [f => (f(succ;succ;zero;@), f(true;@))]}
}

@@

\Par{Path Sensitivity} 
Functions on lists can be generic, in that they may be independent of the item values, as in \code{repeat}.
Additionally, they may factor out the details that depend on item values, as in \code{sort}.

@@

\noindent
Having many different specializations burdens the client of
the generic function to know many different names for roughly the same behavior and can obscure the meaning
of code at the call site. 
Untyped languages or extrinsically typed languages can bypass this rigidity found in 
intrinsically typed languages by mixing the specialized procedures into a single generalized procedures.

\examp{
  \J{def genCmp = [a,b => } 
  \\
  \I \J{a |> (}
  \\
  \I\I \J{[zero;@ => scalarCmp(a,b)]}
  \\
  \I\I \J{[succ;n => scalarCmp(a,b)]}
  \\
  \I\I \J{[nil;@ => lexicoCmp(a,b)]}
  \\
  \I\I \J{[cons;(x,xs) => lexicoCmp(a,b)]}
  \\
  \I \J{)}
  \\
  \J{] in} 
  \\
  \J{...} 
}

\noindent
Extrinsic type inference allows this mixing by inferring types locally for each path,
such that the type inferred from one path is not required for another.
As such, a generalized comparison procedure may be constructed and the
programmer can use the same sort function no matter if its inputs are scalars or vectors.

\examp{
  \J{def genSort : (TOP}
  \\
  \I \J{\& (}\textbf{List}(\textbf{Nat}) \J{ -> } \textbf{List}(\textbf{Nat})
  \\
  \I \J{\& (}\textbf{List}(\textbf{List}(\textbf{nat})) \J{ -> } \textbf{List}(\textbf{List}(\textbf{Nat}))
  \\
  \J{) = sort(genCmp)}
}

\noindent
This particular form of type inference, known as occurrence typing \cite{},
has already been studied and used to great effect in type inference for Python and Javascript.
Additionally, the path sensitivity of type inference obviates the need 
for destruction in pattern matching.  
For example, the \code{repeat} function can be rewritten.

\examp{
  \J{def repeat = [x => loop([self => [n =>}
  \\
  \I\J{n |> [}
  \\
  \I\I \J{[zero;\_ => nil;(n.zero)]}
  \\
  \I\I \J{[succ;\_ => cons;(x,self(n.succ))]}
  \\
  \I \J{]}
  \\
  \J{])]}
}

\noindent
Thus, programs in Python and Javascript, which only having conditional branching,
and no pattern matching, can rely on path sensitivity to simulate pattern matching.

\Par{Pruning}

\TODO{...}

\Par{Streaming}

\TODO{...}

@@

    \inferrule {
      \TODO{modify with case for non-recursive parameter type}
      \\\\
      \Theta, \Delta, \Gamma \entails e \hastype \tau \given  \Theta', \Delta'
      \\
      \Theta', \Delta' \entails \tau \subtypes \alpha \rightarrow \Pi
      \\
      \alpha \notin \textbf{ftv}(\Delta')
      \\\\
      \alpha \downarrow \Pi
      \fallingdotseq 
      \alpha' \uparrow \Pi'
      \\
      \textbf{pack}^-(\textbf{ftv}(\Delta')\ \alpha' \entails \Pi') = \tau'
      \\\\
      \factorsinto \J{LFP[} \alpha' \J{]}\tau' \subtypes \J{left} \mapsto \tau_l
      \\
      \factorsinto \J{LFP[} \alpha' \J{]}\tau' \subtypes \J{right} \mapsto \tau_r
    } {
      \Theta, \Delta, \Gamma \entails \J{loop(}e\J{)} 
      \hastype 
      \tau_l \J{->} \tau_r
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

\begin{definition} 
  \label{def:proof_subtyping_difference_intro}
  \emph{proof subtyping (difference introduction)}
  \hfill
  \small
  \boxed{\Theta, \Delta \entails \tau_l \subtypes \tau_r \typdiff \eta \given \Theta', \Delta'}
  \nopad
  \begin{mathpar}
    \inferrule {
      \Theta, \Delta \Vvdash \eta
      \\
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r \given \Theta', \Delta' 
      \\
      \nexists \Theta', \Delta' \qua  
      \Theta, \Delta \entails 
      \tau_l \subtypes \eta \given \Theta', \Delta'
    } {
      \Theta, \Delta \entails 
      \tau_l \subtypes \tau_r \typdiff \eta
      \given \Theta', \Delta'
    }
  \end{mathpar}
\end{definition}

@@

\begin{requirement}
  \label{req:schema_sequence_duality_correspondence}
  \emph{schema sequence duality correspondence}
  \footnotesize
  \nopad
  \begin{mathpar}
    \inferrule {
      \alpha_\nu \downarrow \Pi_\nu \fallingdotseq \alpha_\mu \uparrow \Pi_\mu
      \\
      \delta \satisfies \Delta
      \\
      \alpha_\nu \notin \textbf{ftv}(\Delta)
      \\
      \alpha_\mu \notin \textbf{ftv}(\Delta)
    } {
      (
      \forall \Theta^\dagger, \Delta^\dagger, \tau^\dagger \qua
      (\Theta^\dagger,\Delta^\dagger, \tau^\dagger) \in \Pi_\nu \implies
      \exists \delta^\dagger \qua \textbf{dom}(\delta^\dagger) \subseteq \Theta^\dagger \up
      \\\\
      (
      \forall \delta' \qua
      \textbf{dom}(\delta') \cap \textbf{ftv}(\Delta) = \epsilon \implies
      % \delta\ \alpha_\nu \slash \J{BOT} \oplus \delta' \oplus \delta^\dagger \satisfies \Delta^\dagger \implies
      % \delta\ \alpha_\nu \slash \J{BOT} \oplus \delta' \oplus \delta^\dagger \satisfies e_f \hastype \tau^\dagger
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies \Delta \cup \Delta^\dagger \implies
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies e_f \hastype \tau^\dagger
      )
      )
      \\\\
      \iff
      \\\\
      (\forall e_a, e_r \qua
      (
      \exists \Theta^\dagger, \Delta^\dagger, \tau^\dagger \qua
      (\Theta^\dagger,\Delta^\dagger, \tau^\dagger) \in \Pi_\mu \up
      (\forall \delta^\dagger \qua \textbf{dom}(\delta^\dagger) \subseteq \Theta^\dagger \implies
      \\\\
      \exists \delta' \qua
      \textbf{dom}(\delta') \cap \textbf{ftv}(\Delta) = \epsilon \up
      % \delta\ \alpha_\mu \slash \J{TOP} \oplus \delta' \oplus \delta^\dagger \satisfies \Delta^\dagger \up
      % \delta\ \alpha_\mu \slash \J{TOP} \oplus \delta' \oplus \delta^\dagger \satisfies (e_a \J{,} e_r) \hastype \tau^\dagger 
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies \Delta \cup \Delta^\dagger \up
      \delta \oplus \delta' \oplus \delta^\dagger \satisfies (e_a \J{,} e_r) \hastype \tau^\dagger 
      )
      )
      \implies
      e_r \cong e_f\J{(}e_a\J{)}
      )
    }
  \end{mathpar}
\end{requirement}

@@

\begin{requirement}
  \label{req:factorization_soundness}
  \emph{factorization soundness}  
  \small
  \begin{mathpar}
    \inferrule {
      \Vdash \J{LFP[}\alpha\J{]}\tau \subtypes l \mapsto \tau'
    } {
      \forall e' \qua
      e' \hastype \tau' 
      \iff 
      (\exists e \qua e \J{.} l \cong e' \up \delta \satisfies e \hastype \J{LFP[}\alpha\J{]}\tau)
    }
  \end{mathpar}
\end{requirement}

@@

    \sdc. &
    \cb{
      \J{[x => halve(double(x))]}
    }
    & \pass & \fail \\
